Project Manager's Progress Report

### 2017-03-12 pnsac/newsletter/article_PresReport.tex
Draft version sent for review.

Since my last Notes from the President in December 2013, restoration
work on the North Star has mostly been suspended. The Museum has
needed extra working space so that exhibits could be made ready for
the commemoration this year of the centenary of the start of the First
World War and the 70th anniversary on June 6 of D-Day in the Second
World War. These preparations have resulted in our normal working
areas being unavailable for North Star restoration work.

\begin{figure}[htbp]
  \vspace{2em}
  \centering
  %name of the graphic, without the path AND in EPS format:
  \includegraphics[scale=0.5]{richard_mtg.eps}
  %caption of the figure 
  \caption*{\small \em Richard at the Quarterly Meeting in April.}
  %label of the figure, which has to correspond to \ref{}:
  \label{fig:richard_mtg}
\end{figure}

During the shut down period some limited work has been continuing. Two
volunteers have been working in the cockpit of the plane and two other
volunteers have been doing some work on \#3 engine reassembly.

While the restoration work has been on hold, we have arranged three
monthly Volunteer Lunches at the Museum to enable the people who
normally work on the plane to meet and socialize. Although we would
all like to be working on the plane, we understand why the Museum has
had to ask us to temporarily suspend our activities. The Volunteer
Lunches are a way of keeping the team together and up-to-date on what
is happening.

At the last meeting of the PNSAC Board of Directors we agreed to
reschedule the Annual General Meeting from June each year to
September. It has always been difficult to find a good weekend in June
to hold the AGM and it is felt that a middle of September date would
be much better. Our corporate secretary, Roger Button, is aware of
this proposed change and has confirmed that the existing Board of
Directors can extend its term until the new AGM date.

Although there has been little restoration work done since December,
we are working with the Museum to get the project back on track again
in early June. We are also working to enable us to recruit new
volunteers to replace the several people who for various reasons have
had to give up active work.

One of the recent discussion areas has been in connection with
long-term funding of the restoration project. Everybody is aware that
funding from the Federal Government has been substantially reduced. We
have accepted that major  expenditures will have to be met from funds
raised by donations from friends of the Museum and members of our
Association. This is a significant challenge since everybody receives
many appeals throughout the year for donations from many different
deserving organizations.

During the next few months we hope to develop a revised restoration
work plan which will enable us to estimate the funds needed to
complete the project. We will then be able to target appeals for funds
to carry out specific parts of the restoration in the same way as we
did for our successful appeal for funds to make troop seats for the
interior of the plane.

>>>>

The last North Star Chronicle was published in March 2016 and since that time we
have successfully continued our work on the aircraft while the Museum was making
preparations for the Star Trek event in the Reserve Hangar. We were not asked to
suspend our work, unlike in 2014 when preparations for the Star Wars event were
being made. This has helped us to retain our Restoration volunteers and recruit
two new ones. However, we are still short of Restoration volunteers and continue
to actively work with the Museum to address this problem.

When Project North Star was originally formed in 2004, the Association was
treated with considerable skepticism by the Museum. It was felt that restoring
an aircraft of the size and complexity of a North Star, to the exacting
standards of a national museum, was likely to be beyond the capabilities of a
group of volunteers and that the project would probably fail.

We will soon have completed 70,000 hours of volunteer work over a period of 14
years, restoring everything we do, to what is acknowledged to be a very high
standard. Skepticism is a thing of the past and we are now treated as a
respected and ongoing part of the Museum operations.

Respect has to be earned and comes through hard work and in the case of aircraft
restoration work, attention to detail. Our team of volunteers needs guidance
both from the Museum staff and from our own senior volunteers. One volunteer has
been particularly important in helping us the get to the position we are now in.
He has recently reached a milestone only reached on two other occasions by a
volunteer and in each case the volunteer worked in the the Science and
Technology Museum. Our volunteer has reached 10,000 volunteer hours of working
on the North Star and throughout this time he also acted as our Project Manager
for everything except the engines. Bruce Gemmill probably was not aware that he
had reached 10,000 hours when he unfortunately had to announce that he would
have to suspend his activities as a volunteer and director of PNSAC due to
family circumstances. We very much hope that he will soon be able to rejoin us
and continue to help us complete this large project. In the meantime Bruce will
continue to be our Membership Secretary. During June, Bruce's remarkable
contribution was recognized by the Museum and in a private event he was
presented with a certificate and a gift by Cedric St Amour, the Museum
Corporation Volunteer Coordinator. One of our volunteers, Robert Desjardins,
also joined Cedric on the occasion to congratulate Bruce on behalf of PNSAC.
Well done Bruce and we are all very proud of your achievement.

I have always seen our Association as having multiple roles. First and foremost
we are restoring a major artifact with our volunteers; secondly we show off our
restoration work to the public which in turn brings visitors and revenue into
the Museum and thirdly through our experienced volunteers we help to transfer
skills to younger and less experienced volunteers.

Planning our work in conjunction with the Museum management and the conservation
staff is something we must always try to jointly do better. It is often a fine
balancing act to decide when to do work on the aircraft in relation to when the
Museum is having a major event such as Canada Day. I am hoping that in future we
can regularly update our work schedule in conjunction with our project plan so
that we can more readily have the aircraft available for public viewing at the
times when the Museum would like us to do so, without having to delay our
restoration work unnecessarily.

I will conclude with an invitation to any of our members not volunteering at
present to come and join us. Although we have a definite limit on the number of
volunteers who can get their hands dirty working on restoration work, we always
welcome others who would like to participate in the running of the organization.
By becoming involved either as a restoration volunteer or an association
volunteer you will meet other like minded people and have an interesting time
talking to members of the public. You do not need to be an expert on aviation or
on the North Star to make a contribution. I can be reached by email or phone if
you would like to find out more about volunteering.

### 2017-03-12 pnsac/newsletter/article_ProgReport.tex
Draft version sent for review.

\title{Project Manager's Progress Report}
\subtitle{November 2015}
\author{Bruce Gemmill}

@@

The aircraft spent considerable time outside this year, but since we
are short of volunteers, not as much work was done as hoped.  The
aircraft is still outside because we are waiting for space to be freed
up in the storage hangar.  We also expect the North Star to be moved
outside again in early spring, much sooner than normal, to accommodate
other uses of the storage facility.  This extended time outside has
delayed some work, and may result in additional weather damage.

@@

The last engine to undergo restoration was disassembled last fall.
Since the engine shop crew have cleaned, painted and re-assembled much
of the engine on the engine assembly stand. The refurbished crankshaft
has been installed in the engine block, and the connecting rods and
pistons cleaned and polished.  Also completed are the more complex
cylinder heads and valve train assemblies, and work has begun to
restore many ancillary pieces, such as the magnetos and other
electrical components, and accessories.

The auxiliary gearbox was removed from Nacelle 4 and has been
completely rebuilt, along with the generator.  These won't be
installed on the aircraft until shortly before engine 4 is ready to be
installed.

Work has also begun on the engine frame which holds the 12 cylinder
engine and supports the surrounding cowl panels.  Once the engine
frame is ready, the engine will be transferred to it from the engine
stand.  Because of a lack of space and shortage of qualified
volunteers, we don't expect to have engine 4 completed before 2017.  A
great deal of work is also needed on the cowl panels.  We hope to
start work on these in the New Year.

@@

Work has continued in the main cabin.  After removing all floor and
wall panels, as well as numerous fittings, the floor was cleaned and
all cargo tiedowns thoroughly treated to remove corrosion.  Several
required removal and repair. The starboard windows were repainted and
new Plexiglas installed, then the windows were re-installed on the
aircraft.  The port windows were then removed and repaired, but had to
be installed prior to completion so the aircraft could stay outside
this summer.  These will need to be removed again this winter and the
new glass installed before being returned to the aircraft.  Some
corrosion damage along the exposed ribs and sub floor has been
completed.  The rear bulkhead required considerable work.  This
involved removing a large corroded floor panel and three rear bulkhead
panels.  New panels were made and are being installed.

The forward bulkhead and forward half of the floor have been painted,
but further painting was stopped by the arrival of cold weather in
October.  The painting will have to be completed next summer.

The forward cargo door was cleaned, painted and new seals installed.
The door was put back on the aircraft to secure it while outside this
summer.  This winter we plan on removing the rear cargo door for
similar work.

The tail section and rear belly compartment underwent some cleaning
and repair.  This work will also continue this winter in the storage
hangar.

\section{Cabin Liners and Troop Seats}
\label{sec:liners_troop_seats}

Most original cabin liners have now been repaired or replaced with new
liners.  The liners still need to be painted and many stencils need to
be applied, as well as many new snap fasteners.

The troop seats were built last year and placed in storage.  Some work
has been done to restore the many fitting needed to install the troop
seats, along with litter straps and tiedowns for litters and cargo.

\section{Planned Restoration Work---2016}
\label{sec:plannedwork}

Over the next year, we will continue to work on Engine 4 and the main
cabin and doors.  The aircraft will be moved outside in the late
winter or early spring to allow setup for a planned Star Trek exhibit.
Work will not begin inside the aircraft until the weather warms up, so
further delays are inevitable.

We will be able to make new wall panels and either replace or restore
the wood floor panels.

We hope to have all main cabin painting done this summer, so that the
wall and floor panels can be installed.

One other work item that may be started is to clean and restore the
four engine nacelles.  Pipes and fittings need to be removed, followed
by a thorough cleaning and a fresh coat of paint.  This work may be
delayed until qualified volunteers become available.

\section{March 2016 Update}
\label{sec:update}

Since our report at the November AGM, the aircraft was moved inside,
so that it could properly sealed against the weather, as it will spend
much of 2016 outside. The aircraft was moved out again in March.
During this time, very little work was done on the aircraft, except
for replacement of the Plexiglas in the port windows, and the
installation of the ARC-552 UHF radio, kindly transferred to the
museum by the Department of National Defence.

\subsection{Nr 4 Engine}
\label{sec:update-engine}

The engine frame has been refurbished, and the engine was transferred
to the frame, so that ancillary equipment can be installed.  Work has
also begun on the supercharger, as well as such items as the starter
motor and header tank.  Some work is also being done on the cowl panels.

%\begin{figure}[htbp]
%   \vspace{2em}
%   \centering
%   %name of the graphic, without the path AND in EPS format:
%   \includegraphics[scale=0.5]{fwd_cockpit_complete.eps}
%   %caption of the figure 
%   \caption*{\small \em Forward cockpit -- complete.}
%   %label of the figure, which has to correspond to \ref{}:
%   \label{fig:fwd_cockpit_complete}
%\end{figure}

>>>>

\title{Project Manager's Progress Report}
\subtitle{PNS 2016 Status Update}
\author{Bruce Gemmill}

@@

@@

Work has continued on much of the ancillary equipment for the engine. The
supercharger and intercooler have been completed and installed, as well as the
reduction gearbox. The radiators and header tank have been restored and painted, and
two of the three radiator sections have been installed. Restoration work has
begun on the air intake system, and many of the fuel and oil lines.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   \includegraphics[scale=0.5]{"PNS 007-im2".png}
   \caption*{\small \em Volunteer Peter Trobridge inspecting the air intake fairing from engine \#4 prior to beginning restoration work.}
   \label{fig:engine_no_4_2}
\end{figure}

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   \includegraphics[scale=0.5]{"PNS 003-im3".png}
   \caption*{\small \em Volunteers Robert Desjardins, Bruce Gemmill and Michel Cote hold the refinished coolant tank from engine \#4.  Hours of polishing were required to get the tank to look like new.}
   \label{fig:coolant_tank}
\end{figure}

The starter motor was found to be very corroded, and disassembly and repair not viable. A
new unit has been ordered and will be installed once delivered. The fire
suppression lines are complete and will be installed along with the cowls when
ready. Some of the cowl panels have been restored, although it has been found
that this engine has more corrosion than the other three engines, possibly it
had been in service longer? This has meant that more work is required to
refurbish each of the many panels. We still hope to have this engine ready for
installation by next spring.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   \includegraphics[scale=0.5]{"PNS 008-im1".png}
   \caption*{\small \em Volunteer Garry Dupont working on the re-assembly of engine \#4.}
   \label{fig:engine_no_4}
\end{figure}

@@

With the aircraft moved outside in March, little work was done in the aircraft
over the winter. Once the temperatures warmed up, we continued work in the main
cabin to prepare for completion of the painting started last year. Repairs to
the rear bulkhead are now complete with the installation of several new pieces
of aluminum to replace corroded ones. A new section of flooring was installed
where the rear washroom was situated. This required replacing about 1000 rivets!

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   \includegraphics[scale=0.5]{"pns 043-im4".png}
   \caption*{\small \em A single new floor panel at the rear of the main cabin required over 1000 rivets.  
   Cleco clips are used to accurately align the panel during riveting.}
   \label{fig:panel_revets}
\end{figure}

Cleaning and corrosion removal continued throughout the cabin. It seems whenever
we feel we are near completion, we find more corrosion. We had also planned to
finish priming the floor and painting the rear bulkhead before the aircraft
returned to the hanger later in the year, but this was delayed by the shutdown
of the project while a new Project Manager was hired by the museum. The wood
floors and walls still need to be repaired, or new panels made, and all will
require painting.

\section{Cabin Liners}
\label{sec:liners}

All the large pieces of liner have either been repaired or replaced, but with
limitations on space, most remaining work on the cabin liners has been
postponed. New window surrounds have also been made, but then all liners will
require painting and stenciling, and this must wait until there is more room in
the shop and the paint booth. Liners were also made for the top of the heater
ducts.

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   \includegraphics[scale=0.5]{P1020369-im5.png}
   \caption*{\small \em One of the refurbished heater duct sections with a new fabric headliner secured to the top. The sides are covered with mica insulation that was removed, cleaned and reattached after repairs to the metal duct. These will be installed in the main cabin.}
   \label{fig:heaters_liners}
\end{figure}

\section{PNS 2017 Update}
\label{sec:pnsupdate}

We were able to restart work on the aircraft in late January, following the
arrival of our new Project manager, Rej Demers. So far, we have completed the
top cowl panels and one of the radiator flaps, which required a repair patch to
the outer skin, and the electrical harnesses, which are now waiting for
painting. Work continues on the air box and the remaining cowl panels.

We have decided not to move the aircraft outside this year. With our efforts
concentrated on completing engine \#4, it is unlikely we would have the resources
to also complete the repair work in the cabin needed to allow for priming and
painting, which is the main reason for moving the aircraft outside in the summer.
